---
title: "EDA"
author: "Fuyu"
date: "5/14/2018"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Project

```{r, echo=FALSE}
author <- function(){
  c('Annie Yang', 'Fuyu Zou', 'Weiqi Pan', 'Lisha Wang')
}
```

***
## 1 Introduction

### 1.1 Import packages
```{r}
library(data.table)
library(ggplot2)
library(leaflet)
library(dplyr)
library(Rmisc)
library(plotly)
library(tidyverse)


```


## 1.2 Load data
```{r}
train = fread('train_taxi.csv')
test = fread('test_taxi.csv')
```

## 1.3 File structure and content
```{r}
summary(train)
str(train)

summary(test)
str(test)
```

We find:

*vender_id* only takes value 1 and 2, we assume it represents two taxi company.
*pickup_datetime* and *dropoff_datetime* are both the combination of date and time, which need to be reformated into sperate variables
*pickup_latitude* and *dropoff_latitude* are not restricted in the New York border (latitude: 40.3371 ~ 41.05555, longitute: -74.4640~-73.4944) as well as *pickup_longitude* and *dropoff_longitude* 
*store_and_fwd_flag* takes two character as 'N' or 'Y'
*trip_duration* which is our target variable, has some extreme value: the min value is 1 second and the maximum value 3526282 second $\approx$ 4 days, which are unbelievable.

## 1.4 missing values

Missing value is a vital part in prediction. If there exists missing value in the dataset, we need to imputate the missing point in order to avoid building model on a small sample. Fortunately, our datasets do not contain any missing value.

```{r}
sum(is.na(train))

sum(is.na(test))
```

# 2 Individual feature visualisations

In this section we will begin by having a look at the distributions of the individual data features.

We start with a map of the pick-up locations of the taxis. The density of the green color shows the frequency of the pick-up coordinates. In this map you can zoom and pan through the pickup locations. 

It is clealy shown on the plot that most pick-up coordinates are whithin Manhattan area. Another two notable hot-spots are LGA and JFK airport towards the south-east of the city.
```{r}
set.seed(1234)
## select a sample of 8000
foo <- sample_n(train, 8e3)


foo %>%
  plot_mapbox(lat = ~pickup_latitude, lon = ~pickup_longitude,
              mode = 'scattermapbox', hoverinfo='speed') %>%
  add_markers(alpha = 0.3, size= I(4), color = I('#32CD32')) %>%
  layout(mapbox = list(style = 'light',
                       zoom = 9,
                       center = list(lat = ~median(pickup_latitude),
                                     lon = ~median(pickup_longitude))))

```

This map shows the drop-off locations. Similar with the pick-up map, most of the cordinates are within Mahattan area,JFK airport and LGA airport,but the locations are more dispersed outside the manhattan area, and more points are located outside the New York city border.

```{r}
foo %>%
  plot_mapbox(lat = ~dropoff_latitude, lon = ~dropoff_longitude,
              mode = 'scattermapbox', hoverinfo='speed') %>%
  add_markers(alpha = 0.3, size= I(4), color = I('#40E0D0')) %>%
  layout(mapbox = list(style = 'light',
                       zoom = 9,
                       center = list(lat = ~median(pickup_latitude),
                                     lon = ~median(pickup_longitude))))
```
On this map, we draw the pickup locations of the two vendors, with orange represents vendor 2 and blue represents verdor 1.

Generally, the two vendors do not have clear their own pick-up location preferences.
```{r}
foo %>%
  plot_mapbox(lat = ~pickup_latitude, lon = ~pickup_longitude, split = ~vendor_id,
              mode = 'scattermapbox', hoverinfo='vendor_id', text =~vendor_id) %>%
  add_markers(alpha = 0.3, size= I(4)) %>%
  layout(mapbox = list(style = 'light',
                       zoom = 9,
                       center = list(lat = ~median(pickup_latitude),
                                     lon = ~median(pickup_longitude))))
```

Now, let's look at the distribution of log10(trip_duration). Note that the y-axis is sqrt of count.

* The majority of rides follow a rather smooth distribution that looks almost log-normal with a peak just short of 1000 seconds, i.e. about 17 minutes.
* There are several suspiciously short rides with less than 10 seconds duration.
* Additionally, there is a strange delta-shaped peak of trip_duration just before the 1e5 seconds mark and even a few way above it:

```{r}

g1 = train %>%
  ggplot(aes(trip_duration)) +
  geom_histogram(bins = 150, fill = 'black', color = 'white') +
  scale_x_log10() +
  scale_y_sqrt() +
  theme_bw()

ggplotly(g1)
```


The distribution of pick-up date and drop-off date

```{r}
###### split date and time 
pickup = stringr::str_split(train$pickup_datetime, ' ')
pick_up = data.frame(matrix(unlist(pickup), ncol = 2, byrow=T))
names(pick_up) = c('pickup_date','pickup_time')
pick_up$pickup_date = as.Date(pick_up$pickup_date,'%Y-%m-%d')
pick_up$pday = format(pick_up$pickup_date,'%m-%d')

dropoff = stringr::str_split(train$dropoff_datetime, ' ')
drop_off = data.frame(matrix(unlist(dropoff), ncol = 2, byrow=T))
names(drop_off) = c('dropoff_date','dropoff_time')
drop_off$dropoff_date = as.Date(drop_off$dropoff_date, '%Y-%m-%d')
drop_off$dday = format(drop_off$dropoff_date, '%m-%d')

train_s = data.frame(cbind(train,pick_up,drop_off)) %>%
  select(-pickup_datetime, -dropoff_datetime)


p1 = plot_ly(train_s, x = ~ pickup_date)  %>% add_histogram(name = "pick_up date")  
p2 = plot_ly(train_s, x = ~ dropoff_date)  %>% add_histogram(name = "drop_off date")  

subplot(p1,p2, nrows = 2, shareX = T,titleX = F )

```

Speed.

```{r}
# compute the direct distiance between pickup and dropoff
pick_coord <- train %>%
  select(pickup_longitude, pickup_latitude)
drop_coord <- train %>%
  select(dropoff_longitude, dropoff_latitude)

# compute the direct distiance between pickup and dropoff
train$dist <- distCosine(pick_coord, drop_coord)

train <- train %>%
  mutate(speed = dist/trip_duration*3.6)


```

```{r}
## plot of direct distance vs trip duration
p_dis_duration = train %>%
  sample_n(5e4) %>%
  ggplot(aes(dist, trip_duration)) +
  geom_point(shape=1, alpha = 0.1) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Direct distance [m]", y = "Trip duration [s]") +
  ggtitle('plot of direct distance vs trip duration') +
  theme_bw()

ggplotly(p_dis_duration)
```



```{r}
## the distribution of average speed
p_speed = train %>%
  filter(speed > 2 & speed < 1e2) %>%
  ggplot(aes(speed)) +
  geom_histogram(bins = 50) +
  labs(x = "Average speed [km/h] (direct distance)") +
  theme_bw()

ggplotly(p_speed)
```


```{r}
# the pickup location with color changing based on average speed
set.seed(1234)
## select a sample of 8000
foo <- sample_n(train, 8e3)

#plotly mapbox
Sys.setenv('MAPBOX_TOKEN' = 'pk.eyJ1IjoiZmZmcmV5YSIsImEiOiJjamg1ZGFrMzIwMHc4MnZwbGU0bDNnaHBlIn0.DFNaPWfAO2wa9wZUNfzLYQ')
foo %>%
  plot_mapbox(lat = ~pickup_latitude, lon = ~pickup_longitude,
              mode = 'scattermapbox', hoverinfo='speed',text = ~speed) %>%
  add_markers(color=~speed, size= I(4), colors = colorRampPalette(c("yellow", "red"))(100)) %>%
  layout(mapbox = list(style = 'light',
                       zoom = 9,
                       center = list(lat = ~median(pickup_latitude),
                                     lon = ~median(pickup_longitude))))
```



```{r}
## the pick-up locations of two vendors
foo %>%
  plot_mapbox(lat = ~pickup_latitude, lon = ~pickup_longitude, split = ~vendor_id,
              mode = 'scattermapbox', hoverinfo='speed',text = ~vendor_id) %>%
  add_markers(alpha = 0.5) %>%
  layout(mapbox = list(style = 'light',
                      zoom = 9,
                      center = list(lat = ~median(pickup_latitude),
                                    lon = ~median(pickup_longitude))))

## the speed comparison of two vendors
train %>%
  group_by(vendor_id) %>%
  plot_ly(y = ~speed, alpha = 0.1) %>%
  add_boxplot(x = ~vendor_id) %>%
  layout(yaxis = list(range= c(0,9285.227)))








```
The summary of average speed of the two vendors contains extreme value. The maximum speed of vendor 2 is 9285 km/h, which is unblievable.